### 2021-03-09 draft2.Rmd
rmd file

---
title: "Movie Recommendation System"
author: "Ray Sherbourne"
date: "March 4, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Project Overview
###section that describes the dataset and summarizes the goal of the project and key steps that were performed

The goal of this project is to solve a popular data science problem known as **The Netflix Challenge**. A dataset of movies and user ratings is provided that will serve as the foundation to build a recommendation system. A validation set is extracted from the data to simulate a true unknown sample and to serve as a final validation test set. The accuracy, as measured against the validation set, is specified as the root mean squared error (RMSE), with a target of 0.86499 or better.

The key steps that were performed are:
1. Import data
  + Download data from source
  + Extract validation set
2. Data wrangling to make data more intuitive
3. Exploratory analysis
4. Model building
  + Start with the mean as a baseline approach
  + Introduce a movie bias term
  + Introduce a user bias term
  + Introduce a regularization term lambda
  + Use matrix factorization
5. Train with best performing model and test on validation set
6. Reflect on the results

#Method
###section that explains the process and techniques used, including data cleaning, data exploration and visualization, insights gained, and your modeling approach

The data was downloaded from the grouplens website and 10% was immidiately set aside as the validation set. Looking at the test set, there appears to be two useful data munging operations we can do. The timestamp predictor appears as a comon format of seconds from xxx. Although this is a common format, it is unintuitive to work with so this is converted to a date time format. The release year may also have some value to our exploration so it is extracted from the movie title.

To start building a model, we will first use the simplest approach to use as a benchmark and itteratively improve upon it. The first model will predict the average movie rating for all movies.

We know that not all movies are recieved by audiences equally.

#Results
###section that presents the modeling results and discusses the model performance

#Conclusion
###section that gives a brief summary of the report, its limitations and future work

### 2021-03-09 draft2.Rmd
added a little bit

@@
   + Introduce a regularization term lambda
-  + Use matrix factorization
+  + Use matrix factorization (with and without optimizing parameters)
 5. Train with best performing model and test on validation set
@@
 
-The data was downloaded from the grouplens website and 10% was immidiately set aside as the validation set.
+The data was downloaded from the grouplens website and 10% was immediately set aside as the validation set.
 Looking at the test set, there appears to be two useful data munging operations we can do.
@@
 
-To start building a model, we will first use the simplest approach to use as a benchmark and itteratively improve upon it. The first model will predict the average movie rating for all movies.
+To start building a model, we will first arrange the data into a matrix that has users in the rows and movies in the columns.
+The intersection between a given movie and a given user is the rating of that movie by that user.
+We will fill in the NA values to predict how each user would rate each movie.
+First we can use the simplest approach as a benchmark and itteratively improve upon it.
+The first model will predict the average movie rating for all movies.
 
-We know that not all movies are recieved by audiences equally.
+We know that not all movies are received by audiences equally. We can account for this by introducing a bias term,
+
+Likewise, different users have different movie preferences and movie critique patterns/tendencies. We can introduce a bias term for users
 

### 2021-03-12 draft2.Rmd
finished outline of #method, ready to add code chunks

@@
 The data was downloaded from the grouplens website and 10% was immediately set aside as the validation set.
-Looking at the test set, there appears to be two useful data munging operations we can do.
-The timestamp predictor appears as a comon format of seconds from xxx.
-Although this is a common format, it is unintuitive to work with so this is converted to a date time format.
+Looking at the test set, there appears to be two useful data munging operations we can do immediately.
+The timestamp predictor appears as a common format of seconds since midnight (UTC) January 1st, 1970.
+Although this is a common format, it is un-intuitive to work with so this is converted to a date time format.
 The release year may also have some value to our exploration so it is extracted from the movie title.
 
-To start building a model, we will first arrange the data into a matrix that has users in the rows and movies in the columns.
-The intersection between a given movie and a given user is the rating of that movie by that user.
-We will fill in the NA values to predict how each user would rate each movie.
-First we can use the simplest approach as a benchmark and itteratively improve upon it.
-The first model will predict the average movie rating for all movies.
+Let's now visualize each of our predictors, aswell as some hypothesis that could merrit further investigation.
 
-We know that not all movies are received by audiences equally. We can account for this by introducing a bias term,
+We will use the simplest approach as a benchmark and itteratively improve upon it. Therefore, our first model will predict the average movie rating for all movies.
 
-Likewise, different users have different movie preferences and movie critique patterns/tendencies. We can introduce a bias term for users
+To account for the fact that movies are not received by audiences equally, we can introducing a bias term.
+This first bias term will represent how a particular movie deviates from the average of all movies.
+
+Likewise, different users have different movie critiquing tendencies. We can introduce a bias term for users that will represent how a particular user deviates from the average user in rating movies.
+
+Now, let's see how the results would change if we reversed the order of the calculation of our bias terms.
+
+Taking the best form of our biased model, we can introduce a regularization term, lambda, to combat over-fitting.
+In the context of the Netflix challenge, over-fitting could occur due to a rarely watched movie being incorrectly labeled as exceptionally good and '`leading to unhelpful recommendations.
+We will use cross-validation to search for the optimal lambda term.
+
+We will now try an alternate approach using matrix factorization.
+The idea behind this approach is that if we arrange the data into a matrix of users x movies, with the intersections representing an individual rating, then we can solve for the matrix of user-preferences and the matrix of movie characteristics, that when multiplied, give the completed matrix which we have just formed.
+Another way of explaining it is that we find similarities between movies and similarities between users.
+Then we can recommend movies that similar users liked, or we can recommend a movie that is similar to others that a user has liked.
+The Recosystem package was built specifically to handle this type of matrix factorization.
+Recosystem takes three vectors: one for the users, one for the movies, and one for the ratings.
+We will include 2 approaches one without tuning the model parameters, and one after tuning the parameters.
+In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
 

### 2021-03-12 main.R
added some exploration

@@ -70,6 +70,37 @@ pattern2 <- "\\d{4}"
 edx <- mutate(edx, release=str_extract(edx$title, pattern))#%>%
 edx <- edx %>% mutate(release=as.numeric(str_extract(edx$release, pattern2)))
 
+# Visualize each of the predictors
+# userId, movieId, rating, timestamp, title, genres, date, release
+
+# User ratings
+edx %>% group_by(userId) %>% summarise(n = n()) %>% 
+  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
+# Movie ratings
+edx %>% group_by(movieId) %>% summarise(n = n()) %>% 
+  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
+# Ratings
+edx %>% group_by(rating) %>% summarise(n = n()) %>%  
+  ggplot(aes(x=rating, y=n)) + geom_line()
+# Timestamp
+edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
+  summarise(rating=mean(rating)) %>%
+  ggplot(aes(x=date, y=rating)) + geom_point() + geom_smooth()
+# Genre vs rating count; as % (see which genre has most ratings)
+# Genre count (see which genre has most movies)
+# Genre rating (see rating distributions of genres {barchart for 1star, 2star, 3.5star filled with %of movies from genre})
+
+
+# Movies by year of release
+# Rating by years after release, grouped by average rating (do good movies get higher ratings with nostalgia and shitty movies start higher because of trailers and hype?)
+
+
+
+
+
+
+
+
 # Christmas movies (proof that holiday movie discrimination is pointless)
 christmas_pattern <- "Christmas"
 christmas_movies <- edx %>% 

### 2021-03-14 draft3.Rmd
updated main and added a second draft of rmd with relevant code chunks

---
title: "Movie Recommendation System"
author: "Ray Sherbourne"
date: "March 13, 2021"
output: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Project Overview
###section that describes the dataset and summarizes the goal of the project and key steps that were performed

The goal of this project is to solve a popular data science problem known as **The Netflix Challenge**. A dataset of movies and user ratings is provided that will serve as the foundation to build a recommendation system. A validation set is extracted from the data to simulate a true unknown sample and to serve as a final validation test set. The accuracy, as measured against the validation set, is specified as the root mean squared error (RMSE), with a target of 0.86499 or better.

The key steps that are to be performed are:
1. Import data
  + Download data from source
  + Extract validation set
2. Data wrangling to clean the data for ease of use
3. Exploratory analysis
4. Model building
  + Start with the mean as a baseline approach
  + Introduce a movie bias term
  + Introduce a user bias term
  + Introduce a regularization term lambda
  + Use matrix factorization (with and without optimizing parameters)
5. Train with best performing model and test on validation set
6. Reflect on the results

#Method
###section that explains the process and techniques used, including data cleaning, data exploration and visualization, insights gained, and your modeling approach

The data was downloaded from the grouplens website and 10% was immediately set aside as the validation set.

```{r, echo=FALSE}
##########################################################
# Create edx set, validation set (final hold-out test set)
##########################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 3.6 or earlier:
#movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
#title = as.character(title),
#genres = as.character(genres))
# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

```

Looking at the test set, there appears to be two useful data munging operations we can do immediately:
* The timestamp predictor appears as a common format of seconds since midnight (UTC) January 1st, 1970. Although this is a common format, it is un-intuitive to work with so this is converted to a date time format. 
```{r, echo=FALSE}
# Converts timestamp to use-able date of review format
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
library(lubridate)
edx <- mutate(edx, date = as_datetime(timestamp))
```
* The release year may have some value to our exploration so it is extracted from the movie title.
```{r, echo=FALSE}
# Extracts release year
if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
library(stringr)
pattern <- "\\(\\d{4}\\)"
pattern2 <- "\\d{4}"
edx <- mutate(edx, release=str_extract(edx$title, pattern))#%>%
edx <- edx %>% mutate(release=as.numeric(str_extract(edx$release, pattern2)))
```

Let's now visualize each of our predictors, to see if there are any apparent patterns or a hypothesis that  merit further investigation.

```{r, echo=FALSE}
# User ratings
edx %>% group_by(userId) %>% summarise(n = n()) %>% 
  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
# Movie ratings
edx %>% group_by(movieId) %>% summarise(n = n()) %>% 
  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
# Ratings
edx %>% group_by(rating) %>% summarise(n = n()) %>%  
  ggplot(aes(x=rating, y=n)) + geom_line()
# Timestamp
edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
  summarise(rating=mean(rating)) %>%
  ggplot(aes(x=date, y=rating)) + geom_point() + geom_smooth()
```

Before we start building our first model, we need to split the data into a training and test set. (Reminder that the validation set replicates data that we don't have yet)
```{r, echo=FALSE}
##########################################################
# Model Building / Select Machine Learning Algorithm
##########################################################

# Split the training set into training/test sets to design/test models
set.seed(420, sample.kind = 'Rounding')
test_index <- createDataPartition(y=edx$rating, times = 1, p=0.1, list=F)
edx_test <- edx[test_index,]
edx_train <- edx[-test_index,]

# This ensures we don't test on movies/users we have never seen before
edx_test <- edx_test %>% semi_join(edx_train, by='movieId') %>%
  semi_join(edx_train, by='userId')
rm(test_index)s
```
We will use the simplest approach as a benchmark and iteratively improve upon it. Therefore, our first model will predict the average movie rating for all movies. 
```{r, echo=FALSE}
# This model predicts the avg movie rating for all cases
mu <- edx_train$rating %>% mean()
(naive_rmse <- RMSE(edx_test$rating, mu))
```
To account for the fact that movies are not received by audiences equally, we can introducing a bias term. This first bias term will represent how a particular movie deviates from the average of all movies.
```{r, echo=FALSE}
# This model builds on previous by introducing a movie bias term
movie_avgs <- edx_train %>% group_by(movieId) %>% summarise(b_i=mean(rating-mu))
movie_bias_model <- mu + edx_test %>% left_join(movie_avgs, by='movieId') %>% pull(b_i)
(movie_bias_rmse <- RMSE(edx_test$rating, movie_bias_model))
```
Likewise, different users have different movie critiquing tendencies. We can introduce a bias term for users that will represent how a particular user deviates from the average user in rating movies.
```{r, echo=FALSE}
# This model builds on previous by introducing a user bias term
user_avgs <- edx_train %>% left_join(movie_avgs, by='movieId') %>% group_by(userId) %>% summarise(b_u=mean(rating-mu-b_i))
user_movie_bias_model <- edx_test %>% left_join(movie_avgs, by='movieId') %>% left_join(user_avgs, by='userId') %>%
  mutate(pred=mu+b_i+b_u) %>% pull(pred)
(user_movie_bias_rmse <- RMSE(edx_test$rating, user_movie_bias_model))
```
Now, let's see how the results would change if we reversed the order of the calculation of our bias terms.
```{r, echo=FALSE}
user_avgs <- edx_train %>% group_by(userId) %>% summarise(b_u=mean(rating-mu))
user_bias_model <- mu + edx_test %>% left_join(user_avgs, by='userId') %>% pull(b_u)
(user_bias_rmse <- RMSE(edx_test$rating, user_bias_model))

# This model builds on previous by introducing a movie bias term
movie_avgs <- edx_train %>% left_join(user_avgs, by='userId') %>% group_by(movieId) %>% summarise(b_i=mean(rating-mu-b_u))
movie_user_bias_model <- edx_test %>% left_join(movie_avgs, by='movieId') %>% left_join(user_avgs, by='userId') %>%
  mutate(pred=mu+b_i+b_u) %>% pull(pred)
(movie_user_bias_rmse <- RMSE(edx_test$rating, movie_user_bias_model))
```
Taking the best form of our biased model, we can introduce a regularization term, lambda, to combat over-fitting. In the context of the Netflix challenge, over-fitting could occur due to a rarely watched movie being incorrectly labeled as exceptionally good and leading to unhelpful recommendations. We will use cross-validation to search for the optimal lambda term.
```{r, echo=FALSE}
# This model builds on biased model by introducing a regularization (lambda) term

# Use cross-validation to search for best lambda term:
#lambdas <- seq(0, 10, 0.25)
lambdas <- seq(4, 5, 0.1)
rmses <- sapply(lambdas, function(l){
  mu <- mean(edx_train$rating)
  b_i <- edx_train %>% group_by(movieId) %>% summarise(b_i=sum(rating-mu)/(n()+l))

  b_u <- edx_train %>% left_join(b_i, by='movieId') %>%
    group_by(userId) %>% summarise(b_u=sum(rating-b_i-mu)/(n()+l))

  predicted_ratings <- edx_test %>% left_join(b_i, by='movieId') %>%
    left_join(b_u, by='userId') %>% mutate(pred=mu+b_i+b_u) %>%
    pull(pred)

  return(RMSE(predicted_ratings, edx_test$rating))
})
# visualize the search for best lambda
qplot(lambdas, rmses)
# save the best lambda term
(lambda <- lambdas[which.min(rmses)])

#Regularized movie bias term
b_i <- edx_train %>% group_by(movieId) %>% summarise(b_i=sum(rating-mu)/(n()+lambda))
#Regularized user bias term
b_u <- edx_train %>% left_join(b_i, by='movieId') %>%
  group_by(userId) %>% summarise(b_u=sum(rating-b_i-mu)/(n()+lambda))

regularized_user_movie_model <- edx_test %>% left_join(b_i, by='movieId') %>%
  left_join(b_u, by='userId') %>% mutate(pred=mu+b_i+b_u) %>% pull(pred)
(regularized_user_movie_rmse <- RMSE(edx_test$rating, regularized_user_movie_model))

```
We will now try an alternate approach using matrix factorization. The idea behind this approach is that if we arrange the data into a matrix of users x movies, with the intersections representing an individual rating, then we can solve for the matrix of user-preferences and the matrix of movie characteristics, that when multiplied, give the completed matrix which we have just formed. Another way of explaining it is that we find similarities between movies and similarities between users. Then we can recommend movies that similar users liked, or we can recommend a movie that is similar to others that a user has liked.
The Recosystem package was built specifically to handle this type of matrix factorization. Recosystem takes three vectors: one for the users, one for the movies, and one for the ratings. We will include 2 approaches one without tuning the model parameters, and one after tuning the parameters. In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
```{r, echo=FALSE}

#This model will use matrix factorization without tuning parameters
if(!require(recosystem)) install.packages("recosystem", repos = "http://cran.us.r-project.org")
set.seed(69, sample.kind = 'Rounding')

train_set <- with(edx_train, data_memory(user_index=userId, item_index=movieId, rating=rating))
test_set <- with(edx_test, data_memory(user_index=userId, item_index=movieId, rating=rating))

# Create a model object
r <- Reco()

#train with tuned parameters
r$train(train_set)

#predict
matrix_factor <- r$predict(test_set, out_memory())

(matrix_factor_rmse <- RMSE(matrix_factor, edx_test$rating))

#This model will tune the training parameters

# Call the $tune() method to select the best parameters with cross validation 
# (This will take a while, see progress bar in R console)
opts <- r$tune(train_set, opts = list(dim=c(10, 20, 30), #dim is number of latent factors
                                      lrate=c(0.1, 0.2), #learning rate (step size in gradient decent)
                                      costp_l1=0, costq_l1=0, #L1 regularization terms set to 0 (default is c(0.01, 0.1))
                                      nthread=1, #number of threads for parallel computing
                                      niter=10)) #number of iterations
opts # $min shows the parameters that minimize the loss function, $res shows the loss function value of each combo

#train with tuned parameters
r$train(train_set, opts=c(opts$min, nthread=1, niter=20))

#predict
matrix_factor_tuned <- r$predict(test_set, out_memory())

(matrix_factor_tuned_rmse <- RMSE(matrix_factor_tuned, edx_test$rating))

```

```{r, echo=FALSE}
# Evaluate the different models
methods <- c('Just the average', '+ movie bias', '+ user bias', 'reversed_biases', 'regularized model', 'matrix factorization', 'matrix factorization (Tuned)')
rmses <- c(naive_rmse, movie_bias_rmse, user_movie_bias_rmse, movie_user_bias_rmse, regularized_user_movie_rmse, matrix_factor_rmse, matrix_factor_tuned_rmse)
(model_evaluations <- tibble(method=methods, RMSE=rmses))
```
#Results
###section that presents the modeling results and discusses the model performance
We now have several models, and we will select the best performing one We will now run our best performing model on the validation set
```{r, echo=FALSE}

```

#Conclusion
###section that gives a brief summary of the report, its limitations and future work

### 2021-03-14 main.R

@@ -60,10 +60,12 @@ rm(dl, ratings, movies, test_index, temp, movielens, removed)
 head(edx)
 
 # Converts timestamp to use-able date of review format
+if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
 library(lubridate)
 edx <- mutate(edx, date = as_datetime(timestamp))
 
 # Extracts release year
+if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
 library(stringr)
 pattern <- "\\(\\d{4}\\)"
 pattern2 <- "\\d{4}"
@@ -86,16 +88,24 @@ edx %>% group_by(rating) %>% summarise(n = n()) %>%
 edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
   summarise(rating=mean(rating)) %>%
   ggplot(aes(x=date, y=rating)) + geom_point() + geom_smooth()
+
+
 # Genre vs rating count; as % (see which genre has most ratings)
+edx %>% group_by(genres) %>%summarise(n=n(), avg=mean(rating), se=sd(rating)/sqrt(n())) %>% mutate(genres=reorder(genres, avg)) %>% ggplot(aes(x = genres, y = avg, ymin = avg - 2*se, ymax = avg + 2*se)) + 
+  geom_point() +
+  geom_errorbar() + 
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
 # Genre count (see which genre has most movies)
+edx %>% group_by(genres) %>% summarise(n=n(), avg=mean(rating)) %>% arrange(desc(n))
 # Genre rating (see rating distributions of genres {barchart for 1star, 2star, 3.5star filled with %of movies from genre})
 
 
 # Movies by year of release
 # Rating by years after release, grouped by average rating (do good movies get higher ratings with nostalgia and shitty movies start higher because of trailers and hype?)
 
-
-
+# Number of movie ratings of movies released in each year
+edx %>% group_by(movieId) %>% summarise(n=n(), year=as.character(release)) %>% 
+  ggplot(aes(year, n)) + geom_boxplot()+ coord_trans(y = "sqrt") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
 
 
 
@@ -122,6 +132,7 @@ edx_train <- edx[-test_index,]
 # This ensures we don't test on movies/users we have never seen before
 edx_test <- edx_test %>% semi_join(edx_train, by='movieId') %>%
   semi_join(edx_train, by='userId')
[…]

### 2021-03-14 draft3.Rmd
r markdown works

@@
 date: "March 13, 2021"
-output: pdf
+output:
+  word_document: default
+  pdf_document: default
 ---
@@
   semi_join(edx_train, by='userId')
-rm(test_index)s
+rm(test_index)
 ```

### 2021-03-15 draft3.Rmd
rmd option hides unwanted messages

@@
 
-```{r, echo=FALSE}
+```{r, include =FALSE}
 ##########################################################
@@
 Although this is a common format, it is un-intuitive to work with so this is converted to a date time format.
-```{r, echo=FALSE}
+```{r, include=FALSE}
 # Converts timestamp to use-able date of review format
@@
 * The release year may have some value to our exploration so it is extracted from the movie title.
-```{r, echo=FALSE}
+```{r, include=FALSE}
 # Extracts release year
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, message=FALSE}
 # User ratings
@@
 Before we start building our first model, we need to split the data into a training and test set. (Reminder that the validation set replicates data that we don't have yet)
-```{r, echo=FALSE}
+```{r, echo=FALSE, warning=FALSE}
 ##########################################################
@@
 This first bias term will represent how a particular movie deviates from the average of all movies.
-```{r, echo=FALSE}
+```{r, echo=FALSE, message=FALSE}
 # This model builds on previous by introducing a movie bias term
@@
 Likewise, different users have different movie critiquing tendencies. We can introduce a bias term for users that will represent how a particular user deviates from the average user in rating movies.
-```{r, echo=FALSE}
+```{r, echo=FALSE, message=FALSE}
 # This model builds on previous by introducing a user bias term
@@
 Now, let's see how the results would change if we reversed the order of the calculation of our bias terms.
-```{r, echo=FALSE}
+```{r, echo=FALSE, message=FALSE}
 user_avgs <- edx_train %>% group_by(userId) %>% summarise(b_u=mean(rating-mu))
@@
 We will use cross-validation to search for the optimal lambda term.
-```{r, echo=FALSE}
+```{r, echo=FALSE, message=FALSE}
 # This model builds on biased model by introducing a regularization (lambda) term
@@
 In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
-```{r, echo=FALSE}
+```{r, echo=FALSE, message=FALSE}
 

### 2021-03-16 draft3.Rmd
building report

@@
 author: "Ray Sherbourne"
-date: "March 13, 2021"
+date: "March 15, 2021"
 output:
@@
 
-#Project Overview
-###section that describes the dataset and summarizes the goal of the project and key steps that were performed
+# Project Overview
 
-The goal of this project is to solve a popular data science problem known as **The Netflix Challenge**.
-A dataset of movies and user ratings is provided that will serve as the foundation to build a recommendation system.
-A validation set is extracted from the data to simulate a true unknown sample and to serve as a final validation test set.
-The accuracy, as measured against the validation set, is specified as the root mean squared error (RMSE), with a target of 0.86499 or better.
+In 2006, Netflix started an open competion to build a model that would predict user ratings for movies using only past ratings as the only predictors.
+This competition has become a popular machine learning challenge in the data science community.
+This report will provide one solution to a modified form of this problem, known as **The Netflix Challenge**, using a dataset provided by grouplens.org.
+The accuracy of the models presented will be measured with the root mean squared error (RMSE), with a target of 0.8775 or lower.
 
-The key steps that are to be performed are:
+The key steps to be performed are:
 1. Import data
@@
 
-#Method
-###section that explains the process and techniques used, including data cleaning, data exploration and visualization, insights gained, and your modeling approach
+# Method
 
-The data was downloaded from the grouplens website and 10% was immediately set aside as the validation set.
+The data was downloaded as a zip file from the grouplens website.
+Two files (movies.dat & ratings.dat) were extracted and joined together by movieId into a dataframe containing columns for userId, movieId, rating, timestamp, title, and genres.
+With the data collection complete, 10% was immediately set aside as the validation set.
 
@@
 
-Looking at the test set, there appears to be two useful data munging operations we can do immediately:
+Two final data munging operations are done on the training set before data exploration:
 * The timestamp predictor appears as a common format of seconds since midnight (UTC) January 1st, 1970.
@@
 
-Let's now visualize each of our predictors, to see if there are any apparent patterns or a hypothesis that  merit further investigation.
+Visual summaries are often useful in conveying and understanding data.
+Simply plotting the predictors will serve as the primary method of data exploration.
+The following histogram shows the number of users that rate a given amount of movies.
 
@@
   ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
+```
+
+The next histogram shows the number of movies that have a given amount of reviews.
+There are a significant amount of movies that have been rated less than 10 times.
+This could potentially be a source of error that may need to be solved with regularization.
+
+```{r, echo=FALSE, message=FALSE}
 # Movie ratings
@@
   ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
+```
+
+The next plot is a line graph showing the frequency of each possible rating. Whole number ratings are much more common.
+
+```{r, echo=FALSE, message=FALSE}
 # Ratings
@@
   ggplot(aes(x=rating, y=n)) + geom_line()
+```
+
+The next plot shows the mean rating for all movies based on the week the review was given. This shows that time-series analysis may have some predictive power, but it is relatively small.
+
+```{r, echo=FALSE, message=FALSE}
 # Timestamp
@@
 
-Before we start building our first model, we need to split the data into a training and test set. (Reminder that the validation set replicates data that we don't have yet)
+After splitting the data into a training and test set (Reminder that the validation set remains untouched), the process of building/testing models begins.
+For this approach, instead of starting with a complex model, a simple model will be built and itteratively improved, while serving as a benchmark to gauge progress.
 ```{r, echo=FALSE, warning=FALSE}
@@
 ```
-We will use the simplest approach as a benchmark and iteratively improve upon it. Therefore, our first model will predict the average movie rating for all movies. 
+The first model will simply predict the average movie rating for all cases:
 ```{r, echo=FALSE}
@@
 ```
-To account for the fact that movies are not received by audiences equally, we can introducing a bias term.
-This first bias term will represent how a particular movie deviates from the average of all movies.
+To improve this model, it is possible to account for the fact that movies are not received by audiences equally by introducing a bias term.
+The bias term is calculated by taking the average of the difference between the individual ratings of a movie and the average of all movie ratings.
+This will represent how a particular movie deviates from the average movie, and will be added onto our previous model.
 ```{r, echo=FALSE, message=FALSE}
@@
 In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
-```{r, echo=FALSE, message=FALSE}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 

### 2021-03-16 draft3.Rmd
pretty tables as output. & finished report sections.

@@
 mu <- edx_train$rating %>% mean()
-(naive_rmse <- RMSE(edx_test$rating, mu))
+naive_rmse <- RMSE(edx_test$rating, mu)
 ```
-To improve this model, it is possible to account for the fact that movies are not received by audiences equally by introducing a bias term.
-The bias term is calculated by taking the average of the difference between the individual ratings of a movie and the average of all movie ratings.
-This will represent how a particular movie deviates from the average movie, and will be added onto our previous model.
+
+Model                                     |    RMSE
+------------------------------------------|-------------
+Average                                   | `naive_rmse`
+
+To improve this model, it is possible to account for the fact that movies are not received by audiences equally.
+A bias term is calculated by taking the average of the difference between the individual ratings of a movie and the average of all movie ratings.
+This will represent how a particular movie deviates from the average movie, and will be added onto our previous model:
 ```{r, echo=FALSE, message=FALSE}
@@
 movie_bias_model <- mu + edx_test %>% left_join(movie_avgs, by='movieId') %>% pull(b_i)
-(movie_bias_rmse <- RMSE(edx_test$rating, movie_bias_model))
+movie_bias_rmse <- RMSE(edx_test$rating, movie_bias_model)
 ```
-Likewise, different users have different movie critiquing tendencies. We can introduce a bias term for users that will represent how a particular user deviates from the average user in rating movies.
+Model                                     |    RMSE
+------------------------------------------|-------------
+Average                                   | `naive_rmse`
+Avg + Movie Bias                          | `movie_bias_rmse`
+
+Likewise, different users have different movie critiquing tendencies.
+A bias term for users, is calculated by taking the average of each users rating after subtracting the average movie rating and movie bias term.
+The user bias term will account for how a particular movie deviates from the average movie for a particular user.
+Again this term is added onto the previous model:
 ```{r, echo=FALSE, message=FALSE}
@@
   mutate(pred=mu+b_i+b_u) %>% pull(pred)
-(user_movie_bias_rmse <- RMSE(edx_test$rating, user_movie_bias_model))
+user_movie_bias_rmse <- RMSE(edx_test$rating, user_movie_bias_model)
 ```
-Now, let's see how the results would change if we reversed the order of the calculation of our bias terms.
+Model                                     |    RMSE
+------------------------------------------|-------------
+Average                                   | `naive_rmse`
+Avg + Movie Bias                          | `movie_bias_rmse`
+Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
+
+The formula for the second bias term included the first bias term, so a quick check is done to see the results of switching the order of calculation:
 ```{r, echo=FALSE, message=FALSE}
@@
 user_bias_model <- mu + edx_test %>% left_join(user_avgs, by='userId') %>% pull(b_u)
-(user_bias_rmse <- RMSE(edx_test$rating, user_bias_model))
+user_bias_rmse <- RMSE(edx_test$rating, user_bias_model)
 
@@
   mutate(pred=mu+b_i+b_u) %>% pull(pred)
-(movie_user_bias_rmse <- RMSE(edx_test$rating, movie_user_bias_model))
+movie_user_bias_rmse <- RMSE(edx_test$rating, movie_user_bias_model)
 ```
-Taking the best form of our biased model, we can introduce a regularization term, lambda, to combat over-fitting.
-In the context of the Netflix challenge, over-fitting could occur due to a rarely watched movie being incorrectly labeled as exceptionally good and leading to unhelpful recommendations.
-We will use cross-validation to search for the optimal lambda term.
+Model                                     |    RMSE
+------------------------------------------|-------------
+Average                                   | `naive_rmse`
+Avg + Movie Bias                          | `movie_bias_rmse`
+Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
+Avg + User Bias                           | `user_bias_rmse`
+Avg + User Bias + Movie Bias              | `movie_user_bias_rmse`
+
+The best form of the model above will be itterated further by adding a regularization term, lambda, to combat over-fitting.
+In the context of The Netflix Challenge, over-fitting could occur due to a rarely watched movie being incorrectly labeled as exceptionally good and leading to unhelpful recommendations.
+Cross-validation is used to search for the optimal lambda term.
 ```{r, echo=FALSE, message=FALSE}
@@
   left_join(b_u, by='userId') %>% mutate(pred=mu+b_i+b_u) %>% pull(pred)
-(regularized_user_movie_rmse <- RMSE(edx_test$rating, regularized_user_movie_model))
+regularized_user_movie_rmse <- RMSE(edx_test$rating, regularized_user_movie_model)
 
 ```
-We will now try an alternate approach using matrix factorization.
-The idea behind this approach is that if we arrange the data into a matrix of users x movies, with the intersections representing an individual rating, then we can solve for the matrix of user-preferences and the matrix of movie characteristics, that when multiplied, give the completed matrix which we have just formed.
-Another way of explaining it is that we find similarities between movies and similarities between users.
-Then we can recommend movies that similar users liked, or we can recommend a movie that is similar to others that a user has liked.
+Model                                     |    RMSE
+------------------------------------------|-------------
+Average                                   | `naive_rmse`
+Avg + Movie Bias                          | `movie_bias_rmse`
+Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
+Avg + User Bias + Movie Bias              | `movie_user_bias_rmse`
+Avg + Movie Bias + User Bias + Lambda     | `regularized_user_movie_rmse`
+
+This model passes the accuracy goal for this project, however there is an alternate approach using matrix factorization.
+The idea behind this approach is that if the data is arranged into a matrix of users x movies, with the intersections representing an individual rating, then factorization would results in a matrix of user-preferences and the matrix of movie characteristics.
+It would be possible to group similar users and group similar movies, and then make recommendations based on these groups.
 The Recosystem package was built specifically to handle this type of matrix factorization.
 Recosystem takes three vectors: one for the users, one for the movies, and one for the ratings.
-We will include 2 approaches one without tuning the model parameters, and one after tuning the parameters.
+This report includes two approaches: one without tuning the model parameters, and one after tuning the parameters.
 In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
@@
 
-```{r, echo=FALSE}
-# Evaluate the different models
-methods <- c('Just the average', '+ movie bias', '+ user bias', 'reversed_biases', 'regularized model', 'matrix factorization', 'matrix factorization (Tuned)')
-rmses <- c(naive_rmse, movie_bias_rmse, user_movie_bias_rmse, movie_user_bias_rmse, regularized_user_movie_rmse, matrix_factor_rmse, matrix_factor_tuned_rmse)
-(model_evaluations <- tibble(method=methods, RMSE=rmses))
-```
+Model                                     |    RMSE
+------------------------------------------|-------------
+Average                                   | `naive_rmse`
+Avg + Movie Bias                          | `movie_bias_rmse`
+Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
+Avg + User Bias + Movie Bias              | `movie_user_bias_rmse`
+Avg + Movie Bias + User Bias + Lambda     | `regularized_user_movie_rmse`
+Matrix Factorization (Untuned)            | `matrix_factor_rmse`
+Matrix Factorization (Tuned)              | `matrix_factor_tuned_rmse`
+
 #Results

### 2021-03-17 draft3.Rmd
updated the draft and finished the main

@@
 
-Model                                     |    RMSE
-------------------------------------------|-------------
-Average                                   | `naive_rmse`
-
 To improve this model, it is possible to account for the fact that movies are not received by audiences equally.
@@
 ```
-Model                                     |    RMSE
-------------------------------------------|-------------
-Average                                   | `naive_rmse`
-Avg + Movie Bias                          | `movie_bias_rmse`
 
@@
 ```
-Model                                     |    RMSE
-------------------------------------------|-------------
-Average                                   | `naive_rmse`
-Avg + Movie Bias                          | `movie_bias_rmse`
-Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
 
@@
 ```
-Model                                     |    RMSE
-------------------------------------------|-------------
-Average                                   | `naive_rmse`
-Avg + Movie Bias                          | `movie_bias_rmse`
-Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
-Avg + User Bias                           | `user_bias_rmse`
-Avg + User Bias + Movie Bias              | `movie_user_bias_rmse`
 
@@
 ```
-Model                                     |    RMSE
-------------------------------------------|-------------
-Average                                   | `naive_rmse`
-Avg + Movie Bias                          | `movie_bias_rmse`
-Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
-Avg + User Bias + Movie Bias              | `movie_user_bias_rmse`
-Avg + Movie Bias + User Bias + Lambda     | `regularized_user_movie_rmse`
 
@@
 
-Model                                     |    RMSE
-------------------------------------------|-------------
-Average                                   | `naive_rmse`
-Avg + Movie Bias                          | `movie_bias_rmse`
-Avg + Movie Bias + User Bias              | `user_movie_bias_rmse`
-Avg + User Bias + Movie Bias              | `movie_user_bias_rmse`
-Avg + Movie Bias + User Bias + Lambda     | `regularized_user_movie_rmse`
-Matrix Factorization (Untuned)            | `matrix_factor_rmse`
-Matrix Factorization (Tuned)              | `matrix_factor_tuned_rmse`
+This method of matrix factorization gives the best result. Even without tuning the parameters, it beats the other models.
 
-#Results
-###section that presents the modeling results and discusses the model performance
-We now have several models, and we will select the best performing one We will now run our best performing model on the validation set
-```{r, echo=FALSE}
+# Results
+### section that presents the modeling results and discusses the model performance
 
+The best performing model is selected to use on the validation set. It is trained with the full edx data set and the rmse of the predictions is calculated below:
+
+```{r}
 ```
 
-#Conclusion
+# Conclusion
 ###section that gives a brief summary of the report, its limitations and future work
+
+In practice the recommendation system would only recommend movies that are predicted to be above the users average rating.
+This validation set replicates how future users would rate movies, both the movies that were recommended and movies that were not recommended (due to a low score).
+
+As the data set is constantly growing, and the algorithm is being tested in the wild, it is realistic to check the rmse of the model post-implementation.
+A data pipeline of importing the newest results and adding them to the existing model would also be important.
+Verifying that the updating rmse is behaving as expected is an easy way of making sure the model has not broken down.

### 2021-03-17 main.R

@@ -72,44 +72,25 @@ pattern2 <- "\\d{4}"
 edx <- mutate(edx, release=str_extract(edx$title, pattern))#%>%
 edx <- edx %>% mutate(release=as.numeric(str_extract(edx$release, pattern2)))
 
-# Visualize each of the predictors
-# userId, movieId, rating, timestamp, title, genres, date, release
+# Visualize the predictors
 
 # User ratings
 edx %>% group_by(userId) %>% summarise(n = n()) %>% 
-  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
+  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')+
+  ggtitle('Users')+ xlab('Number of Movie Ratings') + ylab('Number of Users')
 # Movie ratings
 edx %>% group_by(movieId) %>% summarise(n = n()) %>% 
-  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
+  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')+
+  ggtitle('Ratings')+xlab('Number of Movie Ratings')+ylab('Number of Movies')
 # Ratings
 edx %>% group_by(rating) %>% summarise(n = n()) %>%  
-  ggplot(aes(x=rating, y=n)) + geom_line()
+  ggplot(aes(x=rating, y=n)) + geom_line() +
+  ggtitle('Movie Ratings') + ylab('Number of Ratings')
 # Timestamp
 edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
   summarise(rating=mean(rating)) %>%
-  ggplot(aes(x=date, y=rating)) + geom_point() + geom_smooth()
-
-
-# Genre vs rating count; as % (see which genre has most ratings)
-edx %>% group_by(genres) %>%summarise(n=n(), avg=mean(rating), se=sd(rating)/sqrt(n())) %>% mutate(genres=reorder(genres, avg)) %>% ggplot(aes(x = genres, y = avg, ymin = avg - 2*se, ymax = avg + 2*se)) + 
-  geom_point() +
-  geom_errorbar() + 
-  theme(axis.text.x = element_text(angle = 90, hjust = 1))
-# Genre count (see which genre has most movies)
-edx %>% group_by(genres) %>% summarise(n=n(), avg=mean(rating)) %>% arrange(desc(n))
[…]

### 2021-03-18 main.R
added genre visualization

@@ -72,6 +72,27 @@ pattern2 <- "\\d{4}"
 edx <- mutate(edx, release=str_extract(edx$title, pattern))#%>%
 edx <- edx %>% mutate(release=as.numeric(str_extract(edx$release, pattern2)))
 
+# Creates binary predictor for each genre
+edx <- edx %>% mutate(Action = ifelse(str_detect(.$genres, 'Action'), 1, 0),
+                      Adventure = ifelse(str_detect(.$genres, 'Adventure'), 1, 0),
+                      Animation = ifelse(str_detect(.$genres, 'Animation'), 1, 0),
+                      Children = ifelse(str_detect(.$genres, 'Children'), 1, 0),
+                      Comedy = ifelse(str_detect(.$genres, 'Comedy'), 1, 0),
+                      Crime = ifelse(str_detect(.$genres, 'Crime'), 1, 0),
+                      Docu = ifelse(str_detect(.$genres, 'Documentary'), 1, 0),
+                      Drama = ifelse(str_detect(.$genres, 'Drama'), 1, 0),
+                      Fantasy = ifelse(str_detect(.$genres, 'Fantasy'), 1, 0),
+                      Noir = ifelse(str_detect(.$genres, 'Noir'), 1, 0),
+                      Horror = ifelse(str_detect(.$genres, 'Horror'), 1, 0),
+                      Musical = ifelse(str_detect(.$genres, 'Musical'), 1, 0),
+                      Mystery = ifelse(str_detect(.$genres, 'Mystery'), 1, 0),
+                      Romance = ifelse(str_detect(.$genres, 'Romance'), 1, 0),
+                      SciFi = ifelse(str_detect(.$genres, 'Sci'), 1, 0),
+                      Thriller = ifelse(str_detect(.$genres, 'Thriller'), 1, 0),
+                      War = ifelse(str_detect(.$genres, 'War'), 1, 0),
+                      Western = ifelse(str_detect(.$genres, 'Western'), 1, 0))
+(genre_count <-edx %>% distinct(edx$movieId, .keep_all = TRUE) %>% select(9:26) %>% 
+  colSums()) %>% barplot(las=2, main = 'Movie Genres', ylab = 'Count')
 # Visualize the predictors
 
 # User ratings
@@ -91,6 +112,8 @@ edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
   summarise(rating=mean(rating)) %>%
[…]

### 2021-03-18 draft4.Rmd
?

--- 2021-03-17 draft3.Rmd
@@
 author: "Ray Sherbourne"
-date: "March 15, 2021"
-output:
-  word_document: default
-  pdf_document: default
+date: "March 17, 2021"
+output: word_document
 ---
@@
 
-In 2006, Netflix started an open competion to build a model that would predict user ratings for movies using only past ratings as the only predictors.
-This competition has become a popular machine learning challenge in the data science community.
+In 2006, Netflix started an open competion to build a machine learning model that would predict user ratings for movies using only past ratings as predictors.
 This report will provide one solution to a modified form of this problem, known as **The Netflix Challenge**, using a dataset provided by grouplens.org.
-The accuracy of the models presented will be measured with the root mean squared error (RMSE), with a target of 0.8775 or lower.
+The accuracy of the models presented will be measured against a validation set that is pulled from the dataset before any processing.
+Accuracy will be measured with the root mean squared error (RMSE), targeting a score of 0.8775 or lower.
+The key steps to be performed are:
 
-The key steps to be performed are:
-1. Import data
-  + Download data from source
-  + Extract validation set
+1. Get the data
+      * Download data from source
+      * Combine the data into a single file
+      * Extract the validation set
 2. Data wrangling to clean the data for ease of use
@@
 4. Model building
-  + Start with the mean as a baseline approach
-  + Introduce a movie bias term
-  + Introduce a user bias term
-  + Introduce a regularization term lambda
-  + Use matrix factorization (with and without optimizing parameters)
+      * Start with the mean as a baseline approach
+      * Introduce a movie bias term
+      * Introduce a user bias term
+      * Introduce a regularization term lambda
+      * Use matrix factorization (with and without optimizing parameters)
 5. Train with best performing model and test on validation set
@@
 The data was downloaded as a zip file from the grouplens website.
-Two files (movies.dat & ratings.dat) were extracted and joined together by movieId into a dataframe containing columns for userId, movieId, rating, timestamp, title, and genres.
+Two files (movies.dat & ratings.dat) were extracted and joined together into a dataframe containing columns for userId, movieId, rating, timestamp, title, and genres.
 With the data collection complete, 10% was immediately set aside as the validation set.
 
-```{r, include =FALSE}
-##########################################################
-# Create edx set, validation set (final hold-out test set)
-##########################################################
+From the website/readme file, the data is described as just over 10 million ratings from users who rated at least 20 movies each.
+Users are only represented by their a unique and anonymized id number.
+Ratings were made on a 5-star scale in increments of 1/2.
+The timestamp variable represents seconds form midnight UTC of January 1, 1970.
+The genres variable contains a list of 18 possible genres, separated with pipes.
+Looking at the first few lines of data, two data munging operations will help with further processing:
 
-# Note: this process could take a couple of minutes
+  * Converting the timestamp to a date time format</li>
+  * Extracting the release year from the title</li>
 
-if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
-if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
-if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
-
-library(tidyverse)
-library(caret)
-library(data.table)
-
-# MovieLens 10M dataset:
-# https://grouplens.org/datasets/movielens/10m/
-# http://files.grouplens.org/datasets/movielens/ml-10m.zip
-
-dl <- tempfile()
-download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
-
-ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
-                 col.names = c("userId", "movieId", "rating", "timestamp"))
-
-movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
-colnames(movies) <- c("movieId", "title", "genres")
-
-# if using R 3.6 or earlier:
-#movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
-#title = as.character(title),
-#genres = as.character(genres))
-# if using R 4.0 or later:
-movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
-                                           title = as.character(title),
-                                           genres = as.character(genres))
-
-movielens <- left_join(ratings, movies, by = "movieId")
-
-# Validation set will be 10% of MovieLens data
-set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
-test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
-edx <- movielens[-test_index,]
-temp <- movielens[test_index,]
-
-# Make sure userId and movieId in validation set are also in edx set
-validation <- temp %>% 
-  semi_join(edx, by = "movieId") %>%
-  semi_join(edx, by = "userId")
-
-# Add rows removed from validation set back into edx set
-removed <- anti_join(temp, validation)
-edx <- rbind(edx, removed)
-
-rm(dl, ratings, movies, test_index, temp, movielens, removed)
-
-```
-
-Two final data munging operations are done on the training set before data exploration:
-* The timestamp predictor appears as a common format of seconds since midnight (UTC) January 1st, 1970.
-Although this is a common format, it is un-intuitive to work with so this is converted to a date time format.
-```{r, include=FALSE}
-# Converts timestamp to use-able date of review format
-if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
-library(lubridate)
-edx <- mutate(edx, date = as_datetime(timestamp))
-```
-* The release year may have some value to our exploration so it is extracted from the movie title.
-```{r, include=FALSE}
-# Extracts release year
-if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
-library(stringr)
-pattern <- "\\(\\d{4}\\)"
-pattern2 <- "\\d{4}"
-edx <- mutate(edx, release=str_extract(edx$title, pattern))#%>%
-edx <- edx %>% mutate(release=as.numeric(str_extract(edx$release, pattern2)))
-```
-
-Visual summaries are often useful in conveying and understanding data.
-Simply plotting the predictors will serve as the primary method of data exploration.
+Visual summaries are often useful in both conveying and understanding data.
+A few simple plots will serve as the primary method of data exploration.
 The following histogram shows the number of users that rate a given amount of movies.
-
-```{r, echo=FALSE, message=FALSE}
-# User ratings
-edx %>% group_by(userId) %>% summarise(n = n()) %>% 
-  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
-```
 
@@
 
-```{r, echo=FALSE, message=FALSE}
-# Movie ratings
-edx %>% group_by(movieId) %>% summarise(n = n()) %>% 
-  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')
-```
-
 The next plot is a line graph showing the frequency of each possible rating. Whole number ratings are much more common.
-
-```{r, echo=FALSE, message=FALSE}
-# Ratings
-edx %>% group_by(rating) %>% summarise(n = n()) %>%  
-  ggplot(aes(x=rating, y=n)) + geom_line()
-```
 
@@
 
-```{r, echo=FALSE, message=FALSE}
-# Timestamp
-edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
-  summarise(rating=mean(rating)) %>%
-  ggplot(aes(x=date, y=rating)) + geom_point() + geom_smooth()
-```
-
 After splitting the data into a training and test set (Reminder that the validation set remains untouched), the process of building/testing models begins.
 For this approach, instead of starting with a complex model, a simple model will be built and itteratively improved, while serving as a benchmark to gauge progress.
-```{r, echo=FALSE, warning=FALSE}
-##########################################################
-# Model Building / Select Machine Learning Algorithm
-##########################################################
 
-# Split the training set into training/test sets to design/test models
-set.seed(420, sample.kind = 'Rounding')
-test_index <- createDataPartition(y=edx$rating, times = 1, p=0.1, list=F)
-edx_test <- edx[test_index,]
-edx_train <- edx[-test_index,]
-
-# This ensures we don't test on movies/users we have never seen before
-edx_test <- edx_test %>% semi_join(edx_train, by='movieId') %>%
-  semi_join(edx_train, by='userId')
-rm(test_index)
-```
 The first model will simply predict the average movie rating for all cases:
-```{r, echo=FALSE}
-# This model predicts the avg movie rating for all cases
-mu <- edx_train$rating %>% mean()
-naive_rmse <- RMSE(edx_test$rating, mu)
-```
 
@@
 This will represent how a particular movie deviates from the average movie, and will be added onto our previous model:
-```{r, echo=FALSE, message=FALSE}
-# This model builds on previous by introducing a movie bias term
-movie_avgs <- edx_train %>% group_by(movieId) %>% summarise(b_i=mean(rating-mu))
-movie_bias_model <- mu + edx_test %>% left_join(movie_avgs, by='movieId') %>% pull(b_i)
-movie_bias_rmse <- RMSE(edx_test$rating, movie_bias_model)
-```
 
@@
 Again this term is added onto the previous model:
-```{r, echo=FALSE, message=FALSE}
-# This model builds on previous by introducing a user bias term
-user_avgs <- edx_train %>% left_join(movie_avgs, by='movieId') %>% group_by(userId) %>% summarise(b_u=mean(rating-mu-b_i))
-user_movie_bias_model <- edx_test %>% left_join(movie_avgs, by='movieId') %>% left_join(user_avgs, by='userId') %>%
-  mutate(pred=mu+b_i+b_u) %>% pull(pred)
-user_movie_bias_rmse <- RMSE(edx_test$rating, user_movie_bias_model)
-```
 
 The formula for the second bias term included the first bias term, so a quick check is done to see the results of switching the order of calculation:
-```{r, echo=FALSE, message=FALSE}
-user_avgs <- edx_train %>% group_by(userId) %>% summarise(b_u=mean(rating-mu))
-user_bias_model <- mu + edx_test %>% left_join(user_avgs, by='userId') %>% pull(b_u)
-user_bias_rmse <- RMSE(edx_test$rating, user_bias_model)
-
-# This model builds on previous by introducing a movie bias term
-movie_avgs <- edx_train %>% left_join(user_avgs, by='userId') %>% group_by(movieId) %>% summarise(b_i=mean(rating-mu-b_u))
-movie_user_bias_model <- edx_test %>% left_join(movie_avgs, by='movieId') %>% left_join(user_avgs, by='userId') %>%
-  mutate(pred=mu+b_i+b_u) %>% pull(pred)
-movie_user_bias_rmse <- RMSE(edx_test$rating, movie_user_bias_model)
-```
 
@@
 Cross-validation is used to search for the optimal lambda term.
-```{r, echo=FALSE, message=FALSE}
-# This model builds on biased model by introducing a regularization (lambda) term
-
-# Use cross-validation to search for best lambda term:
-#lambdas <- seq(0, 10, 0.25)
-lambdas <- seq(4, 5, 0.1)
-rmses <- sapply(lambdas, function(l){
-  mu <- mean(edx_train$rating)
-  b_i <- edx_train %>% group_by(movieId) %>% summarise(b_i=sum(rating-mu)/(n()+l))
-
-  b_u <- edx_train %>% left_join(b_i, by='movieId') %>%
-    group_by(userId) %>% summarise(b_u=sum(rating-b_i-mu)/(n()+l))
-
-  predicted_ratings <- edx_test %>% left_join(b_i, by='movieId') %>%
-    left_join(b_u, by='userId') %>% mutate(pred=mu+b_i+b_u) %>%
-    pull(pred)
-
-  return(RMSE(predicted_ratings, edx_test$rating))
-})
-# visualize the search for best lambda
-qplot(lambdas, rmses)
-# save the best lambda term
-(lambda <- lambdas[which.min(rmses)])
-
-#Regularized movie bias term
-b_i <- edx_train %>% group_by(movieId) %>% summarise(b_i=sum(rating-mu)/(n()+lambda))
-#Regularized user bias term
-b_u <- edx_train %>% left_join(b_i, by='movieId') %>%
-  group_by(userId) %>% summarise(b_u=sum(rating-b_i-mu)/(n()+lambda))
-
-regularized_user_movie_model <- edx_test %>% left_join(b_i, by='movieId') %>%
-  left_join(b_u, by='userId') %>% mutate(pred=mu+b_i+b_u) %>% pull(pred)
-regularized_user_movie_rmse <- RMSE(edx_test$rating, regularized_user_movie_model)
-
-```
 
 This model passes the accuracy goal for this project, however there is an alternate approach using matrix factorization.
-The idea behind this approach is that if the data is arranged into a matrix of users x movies, with the intersections representing an individual rating, then factorization would results in a matrix of user-preferences and the matrix of movie characteristics.
+The idea behind this approach is that if the data is arranged into a matrix of users x movies, with the intersections representing an individual rating, then factorization would results in a matrix of user-preferences and a matrix of movie characteristics.
 It would be possible to group similar users and group similar movies, and then make recommendations based on these groups.
@@
 In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-
-#This model will use matrix factorization without tuning parameters
-if(!require(recosystem)) install.packages("recosystem", repos = "http://cran.us.r-project.org")
-set.seed(69, sample.kind = 'Rounding')
-
-train_set <- with(edx_train, data_memory(user_index=userId, item_index=movieId, rating=rating))
-test_set <- with(edx_test, data_memory(user_index=userId, item_index=movieId, rating=rating))
-
-# Create a model object
-r <- Reco()
-
-#train with tuned parameters
-r$train(train_set)
-
-#predict
-matrix_factor <- r$predict(test_set, out_memory())
-
-(matrix_factor_rmse <- RMSE(matrix_factor, edx_test$rating))
-
-#This model will tune the training parameters
-
-# Call the $tune() method to select the best parameters with cross validation 
-# (This will take a while, see progress bar in R console)
-opts <- r$tune(train_set, opts = list(dim=c(10, 20, 30), #dim is number of latent factors
-                                      lrate=c(0.1, 0.2), #learning rate (step size in gradient decent)
-                                      costp_l1=0, costq_l1=0, #L1 regularization terms set to 0 (default is c(0.01, 0.1))
-                                      nthread=1, #number of threads for parallel computing
-                                      niter=10)) #number of iterations
-opts # $min shows the parameters that minimize the loss function, $res shows the loss function value of each combo
-
-#train with tuned parameters
-r$train(train_set, opts=c(opts$min, nthread=1, niter=20))
-
-#predict
-matrix_factor_tuned <- r$predict(test_set, out_memory())
-
-(matrix_factor_tuned_rmse <- RMSE(matrix_factor_tuned, edx_test$rating))
-
-```
 
@@
 # Results
-### section that presents the modeling results and discusses the model performance
 
@@
 
-```{r}
-```
+# Conclusion
 
-# Conclusion
-###section that gives a brief summary of the report, its limitations and future work
+The goal of this report was to build a movie recomendation system using a dataset of past reviews.
+After importation, the data was cleaned and briefly explored.
+During the exploration phase, some potential patterns/problems in the data were identified.
+Several models were built using the insights from the explorations and prior intuition.
+In the end the goal of an rmse score below 0.8775 was achieved, but in practice the effort to improve the models would continue.
+One such further improvement could be implementing the genres feature into the model.
+Some of the limitations of the models presented here are:
 
-In practice the recommendation system would only recommend movies that are predicted to be above the users average rating.
-This validation set replicates how future users would rate movies, both the movies that were recommended and movies that were not recommended (due to a low score).
+  * It is unclear how many users may be sharing a single id. Adding a feature on a movie streaming platform to account for multiple users would result in more personalized predictions.
+  * The data for a users search history is a powerfull predictor that a service like Netflix would have access to.
+  * The given database does not include features such as actors and directors due to limitations of the authors computer.
+  * The total amount of time a user spends watching a movie, could have some predictive power for other users. For example: multiple re-watches may have more value than a 5-star rating.
 
 As the data set is constantly growing, and the algorithm is being tested in the wild, it is realistic to check the rmse of the model post-implementation.
-A data pipeline of importing the newest results and adding them to the existing model would also be important.
+A data pipeline of importing the newest results and adding them to the existing model would be important.
 Verifying that the updating rmse is behaving as expected is an easy way of making sure the model has not broken down.

### 2021-03-21 draft4.Rmd
Updated rmd

@@
 author: "Ray Sherbourne"
-date: "March 17, 2021"
+date: "March 21, 2021"
 output: word_document
@@
 
-In 2006, Netflix started an open competion to build a machine learning model that would predict user ratings for movies using only past ratings as predictors.
-This report will provide one solution to a modified form of this problem, known as **The Netflix Challenge**, using a dataset provided by grouplens.org.
-The accuracy of the models presented will be measured against a validation set that is pulled from the dataset before any processing.
+In 2006, Netflix started an open competition to build a machine learning model that would predict user ratings for movies using only past ratings as predictors.
+This report will provide one solution to a modified form of this problem, known as **The Netflix Prize**, using a data set provided by grouplens.org.
+The accuracies of the models presented will be measured against a validation set that is pulled from the original data set before any processing.
 Accuracy will be measured with the root mean squared error (RMSE), targeting a score of 0.8775 or lower.
@@
       * Extract the validation set
-2. Data wrangling to clean the data for ease of use
+2. Data wrangling (clean the data for ease of use)
 3. Exploratory analysis
@@
       * Start with the mean as a baseline approach
-      * Introduce a movie bias term
-      * Introduce a user bias term
+      * Account for movie bias
+      * Account for user bias
       * Introduce a regularization term lambda
@@
 
+```{r, include =FALSE}
+##########################################################
+# Create edx set, validation set (final hold-out test set)
+##########################################################
+
+# Note: this process could take a couple of minutes
+
+if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
+if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
+if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
+
+library(tidyverse)
+library(caret)
+library(data.table)
+
+# MovieLens 10M dataset:
+# https://grouplens.org/datasets/movielens/10m/
+# http://files.grouplens.org/datasets/movielens/ml-10m.zip
+
+dl <- tempfile()
+download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
+
+ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
+                 col.names = c("userId", "movieId", "rating", "timestamp"))
+
+movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
+colnames(movies) <- c("movieId", "title", "genres")
+
+# if using R 3.6 or earlier:
+#movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
+#title = as.character(title),
+#genres = as.character(genres))
+# if using R 4.0 or later:
+movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
+                                           title = as.character(title),
+                                           genres = as.character(genres))
+
+movielens <- left_join(ratings, movies, by = "movieId")
+
+# Validation set will be 10% of MovieLens data
+set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
+test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
+edx <- movielens[-test_index,]
+temp <- movielens[test_index,]
+
+# Make sure userId and movieId in validation set are also in edx set
+validation <- temp %>% 
+  semi_join(edx, by = "movieId") %>%
+  semi_join(edx, by = "userId")
+
+# Add rows removed from validation set back into edx set
+removed <- anti_join(temp, validation)
+edx <- rbind(edx, removed)
+
+rm(dl, ratings, movies, test_index, temp, movielens, removed)
+```
+
 From the website/readme file, the data is described as just over 10 million ratings from users who rated at least 20 movies each.
 Users are only represented by their a unique and anonymized id number.
-Ratings were made on a 5-star scale in increments of 1/2.
+Ratings were made on a scale from 0.5 stars (worst) to 5 stars (best) in increments of 0.5.
 The timestamp variable represents seconds form midnight UTC of January 1, 1970.
-The genres variable contains a list of 18 possible genres, separated with pipes.
-Looking at the first few lines of data, two data munging operations will help with further processing:
+The genres variable contains a combinations of 18 possible genres.
+Looking at the first few lines of data, there are a few data munging operations that will help with further processing:
 
-  * Converting the timestamp to a date time format</li>
-  * Extracting the release year from the title</li>
+```{r, echo=FALSE}
+head(edx)
+```
+
+  * Converting the timestamp to a date time format
+  * Extracting the release year from the title
+  * Pulling out the genres into their own separate binary columns
+
+```{r, include=FALSE}
+# Converts timestamp to use-able date of review format
+if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
+library(lubridate)
+edx <- mutate(edx, date = as_datetime(timestamp))
+
+# Extracts release year
+if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
+library(stringr)
+pattern <- "\\(\\d{4}\\)"
+pattern2 <- "\\d{4}"
+edx <- mutate(edx, release=str_extract(edx$title, pattern))#%>%
+edx <- edx %>% mutate(release=as.numeric(str_extract(edx$release, pattern2)))
+
+# Creates binary predictor for each genre
+edx <- edx %>% mutate(Action = ifelse(str_detect(.$genres, 'Action'), 1, 0),
+                      Adventure = ifelse(str_detect(.$genres, 'Adventure'), 1, 0),
+                      Animation = ifelse(str_detect(.$genres, 'Animation'), 1, 0),
+                      Children = ifelse(str_detect(.$genres, 'Children'), 1, 0),
+                      Comedy = ifelse(str_detect(.$genres, 'Comedy'), 1, 0),
+                      Crime = ifelse(str_detect(.$genres, 'Crime'), 1, 0),
+                      Docu = ifelse(str_detect(.$genres, 'Documentary'), 1, 0),
+                      Drama = ifelse(str_detect(.$genres, 'Drama'), 1, 0),
+                      Fantasy = ifelse(str_detect(.$genres, 'Fantasy'), 1, 0),
+                      Noir = ifelse(str_detect(.$genres, 'Noir'), 1, 0),
+                      Horror = ifelse(str_detect(.$genres, 'Horror'), 1, 0),
+                      Musical = ifelse(str_detect(.$genres, 'Musical'), 1, 0),
+                      Mystery = ifelse(str_detect(.$genres, 'Mystery'), 1, 0),
+                      Romance = ifelse(str_detect(.$genres, 'Romance'), 1, 0),
+                      SciFi = ifelse(str_detect(.$genres, 'Sci'), 1, 0),
+                      Thriller = ifelse(str_detect(.$genres, 'Thriller'), 1, 0),
+                      War = ifelse(str_detect(.$genres, 'War'), 1, 0),
+                      Western = ifelse(str_detect(.$genres, 'Western'), 1, 0))
+(genre_count <-edx %>% distinct(edx$movieId, .keep_all = TRUE) %>% select(9:26) %>% 
+  colSums()) %>% barplot(las=2, main = 'Movie Genres', ylab = 'Count')
+```
 
 Visual summaries are often useful in both conveying and understanding data.
-A few simple plots will serve as the primary method of data exploration.
+A few simple plots will serve as the primary method of data exploration here.
 The following histogram shows the number of users that rate a given amount of movies.
+```{r, echo=FALSE, message=FALSE}
+# User ratings
+edx %>% group_by(userId) %>% summarise(n = n()) %>% 
+  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')+
+  ggtitle('Users')+ xlab('Number of Movie Ratings') + ylab('Number of Users')
+```
 
@@
 
-The next plot is a line graph showing the frequency of each possible rating. Whole number ratings are much more common.
+```{r, echo=FALSE, message=FALSE}
+# Movie ratings
+edx %>% group_by(movieId) %>% summarise(n = n()) %>% 
+  ggplot(aes(n)) + geom_histogram() + scale_x_continuous(trans='log10')+
+  ggtitle('Ratings')+xlab('Number of Movie Ratings')+ylab('Number of Movies')
+```
+
+The next plot is a line graph showing the frequency of each possible rating. It is obvious that whole number ratings are much more common.
+
+```{r, echo=FALSE, message=FALSE}
+# Ratings
+edx %>% group_by(rating) %>% summarise(n = n()) %>%  
+  ggplot(aes(x=rating, y=n)) + geom_line() +
+  ggtitle('Movie Ratings') + ylab('Number of Ratings')
+```
 
@@
 
-After splitting the data into a training and test set (Reminder that the validation set remains untouched), the process of building/testing models begins.
+```{r, echo=FALSE, message=FALSE}
+# Timestamp
+edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
+  summarise(rating=mean(rating)) %>%
+  ggplot(aes(x=date, y=rating)) + geom_point() + geom_smooth()+
+  ggtitle('Average Rating per week') + ylab('Rating')
+```
+
+To follow up on the previous idea of using time as a predictor.
+A check is done using movies containing the word 'Christmas' to see if holiday themed movies should be recommended more around that holiday.
+
+```{r, echo=FALSE, message=FALSE}
+# Christmas movies (proof that holiday movie discrimination is pointless)
+christmas_pattern <- "Christmas"
+christmas_movies <- edx %>% 
+  mutate(xmas = ifelse(str_detect(edx$title, christmas_pattern), 1, 0))
+christmas_movies <- christmas_movies[christmas_movies$xmas==1] %>% group_by(title)
+
+christmas_movies %>% ggplot(aes(x=month(date))) + geom_bar(stat = 'count') +
+  scale_x_continuous(breaks=c(1:12)) +
+  ggtitle("Reviews for Christmas Movies")+
+  xlab('Month')
+```
+
+These results may be somewhat surprising and suggest that holiday movies will not impact a recommendation system in a meaningful way.
+
+Using the insights from the readme file, and the exploratory analysis, prediction models are ready to be built.
+But first the data is split into separate training and test sets, so that the model accuracy can be measured (Reminder that the validation set remains untouched).
 For this approach, instead of starting with a complex model, a simple model will be built and itteratively improved, while serving as a benchmark to gauge progress.
 
+```{r, include=FALSE, warning=FALSE}
+# Split the training set into training/test sets to design/test models
+set.seed(420, sample.kind = 'Rounding')
+test_index <- createDataPartition(y=edx$rating, times = 1, p=0.1, list=F)
+edx_test <- edx[test_index,]
+edx_train <- edx[-test_index,]
+
+# This ensures we don't test on movies/users we have never seen before
+edx_test <- edx_test %>% semi_join(edx_train, by='movieId') %>%
+  semi_join(edx_train, by='userId')
+rm(test_index)
+```
+
 The first model will simply predict the average movie rating for all cases:
+
+```{r, echo=FALSE, message=FALSE}
+# This model predicts the avg movie rating for all cases
+mu <- edx_train$rating %>% mean()
+naive_rmse <- RMSE(edx_test$rating, mu)
+
+methods <- ('Just the average')
+rmses <- (naive_rmse)
+(model_evaluations <- tibble(method=methods, RMSE=rmses))
+```
 
@@
 
+```{r, echo=FALSE, message=FALSE}
+# This model builds on previous by introducing a movie bias term
+movie_avgs <- edx_train %>% group_by(movieId) %>% summarise(b_i=mean(rating-mu))
+movie_bias_model <- mu + edx_test %>% left_join(movie_avgs, by='movieId') %>% pull(b_i)
+movie_bias_rmse <- RMSE(edx_test$rating, movie_bias_model)
+
+methods <- c('Just the average', '+ movie bias')
+rmses <- c(naive_rmse, movie_bias_rmse)
+(model_evaluations <- tibble(method=methods, RMSE=rmses))
+```
 Likewise, different users have different movie critiquing tendencies.
@@
 Again this term is added onto the previous model:
+```{r, echo=FALSE, message=FALSE}
+# This model builds on previous by introducing a user bias term
+user_avgs <- edx_train %>% left_join(movie_avgs, by='movieId') %>% group_by(userId) %>% summarise(b_u=mean(rating-mu-b_i))
+user_movie_bias_model <- edx_test %>% left_join(movie_avgs, by='movieId') %>% left_join(user_avgs, by='userId') %>%
+  mutate(pred=mu+b_i+b_u) %>% pull(pred)
+user_movie_bias_rmse <- RMSE(edx_test$rating, user_movie_bias_model)
+
+methods <- c('Just the average', '+ movie bias', '+ user bias')
+rmses <- c(naive_rmse, movie_bias_rmse, user_movie_bias_rmse)
+(model_evaluations <- tibble(method=methods, RMSE=rmses))
+```
 
 The formula for the second bias term included the first bias term, so a quick check is done to see the results of switching the order of calculation:
+
+```{r, echo=FALSE, message=FALSE}
+# This model builds on mu by introducing a user bias term
+user_avgs <- edx_train %>% group_by(userId) %>% summarise(b_u=mean(rating-mu))
+user_bias_model <- mu + edx_test %>% left_join(user_avgs, by='userId') %>% pull(b_u)
+user_bias_rmse <- RMSE(edx_test$rating, user_bias_model)
+
+# This model builds on previous by introducing a movie bias term
+movie_avgs <- edx_train %>% left_join(user_avgs, by='userId') %>% group_by(movieId) %>% summarise(b_i=mean(rating-mu-b_u))
+movie_user_bias_model <- edx_test %>% left_join(movie_avgs, by='movieId') %>% left_join(user_avgs, by='userId') %>%
+  mutate(pred=mu+b_i+b_u) %>% pull(pred)
+movie_user_bias_rmse <- RMSE(edx_test$rating, movie_user_bias_model)
+
+methods <- c('Just the average', '+ movie bias', '+ user bias', 'reversed_biases')
+rmses <- c(naive_rmse, movie_bias_rmse, user_movie_bias_rmse, movie_user_bias_rmse)
+(model_evaluations <- tibble(method=methods, RMSE=rmses))
+```
 
@@
 Cross-validation is used to search for the optimal lambda term.
+
+```{r, echo=FALSE, message=FALSE}
+# This model builds on biased model by introducing a regularization (lambda) term
+
+# Use cross-validation to search for best lambda term:
+#lambdas <- seq(0, 10, 0.25)
+lambdas <- seq(4, 5, 0.1)
+rmses <- sapply(lambdas, function(l){
+  mu <- mean(edx_train$rating)
+  b_i <- edx_train %>% group_by(movieId) %>% summarise(b_i=sum(rating-mu)/(n()+l))
+
+  b_u <- edx_train %>% left_join(b_i, by='movieId') %>%
+    group_by(userId) %>% summarise(b_u=sum(rating-b_i-mu)/(n()+l))
+
+  predicted_ratings <- edx_test %>% left_join(b_i, by='movieId') %>%
+    left_join(b_u, by='userId') %>% mutate(pred=mu+b_i+b_u) %>%
+    pull(pred)
+
+  return(RMSE(predicted_ratings, edx_test$rating))
+})
+# visualize the search for best lambda
+qplot(lambdas, rmses)
+# save the best lambda term
+(lambda <- lambdas[which.min(rmses)])
+
+#Regularized movie bias term
+b_i <- edx_train %>% group_by(movieId) %>% summarise(b_i=sum(rating-mu)/(n()+lambda))
+#Regularized user bias term
+b_u <- edx_train %>% left_join(b_i, by='movieId') %>%
+  group_by(userId) %>% summarise(b_u=sum(rating-b_i-mu)/(n()+lambda))
+
+regularized_user_movie_model <- edx_test %>% left_join(b_i, by='movieId') %>%
+  left_join(b_u, by='userId') %>% mutate(pred=mu+b_i+b_u) %>% pull(pred)
+regularized_user_movie_rmse <- RMSE(edx_test$rating, regularized_user_movie_model)
+
+methods <- c('Just the average', '+ movie bias', '+ user bias', 'reversed_biases', 'regularized model')
+rmses <- c(naive_rmse, movie_bias_rmse, user_movie_bias_rmse, movie_user_bias_rmse, regularized_user_movie_rmse)
+(model_evaluations <- tibble(method=methods, RMSE=rmses))
+```
 
@@
 In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+#This model will use matrix factorization without tuning parameters
+if(!require(recosystem)) install.packages("recosystem", repos = "http://cran.us.r-project.org")
+set.seed(69, sample.kind = 'Rounding')
+
+train_set <- with(edx_train, data_memory(user_index=userId, item_index=movieId, rating=rating))
+test_set <- with(edx_test, data_memory(user_index=userId, item_index=movieId, rating=rating))
+
+# Create a model object
+r <- Reco()
+
+#train with tuned parameters
+r$train(train_set)
+
+#predict
+matrix_factor <- r$predict(test_set, out_memory())
+
+matrix_factor_rmse <- RMSE(matrix_factor, edx_test$rating)
+
+#This model will tune the training parameters
+
+# Call the $tune() method to select the best parameters with cross validation 
+# (This will take a while, see progress bar in R console)
+opts <- r$tune(train_set, opts = list(dim=c(10, 20, 30), #dim is number of latent factors
+                                      lrate=c(0.1, 0.2), #learning rate (step size in gradient decent)
+                                      costp_l1=0, costq_l1=0, #L1 regularization terms set to 0 (default is c(0.01, 0.1))
+                                      nthread=1, #number of threads for parallel computing
+                                      niter=10)) #number of iterations
+
+#train with tuned parameters
+r$train(train_set, opts=c(opts$min, nthread=1, niter=20))
+
+#predict
+matrix_factor_tuned <- r$predict(test_set, out_memory())
+
+matrix_factor_tuned_rmse <- RMSE(matrix_factor_tuned, edx_test$rating)
+
+# Evaluate the different models
+methods <- c('Just the average', '+ movie bias', '+ user bias', 'reversed_biases', 'regularized model', 'matrix factorization', 'matrix factorization (Tuned)')
+rmses <- c(naive_rmse, movie_bias_rmse, user_movie_bias_rmse, movie_user_bias_rmse, regularized_user_movie_rmse, matrix_factor_rmse, matrix_factor_tuned_rmse)
+(model_evaluations <- tibble(method=methods, RMSE=rmses))
+
+```
 
@@
 The best performing model is selected to use on the validation set. It is trained with the full edx data set and the rmse of the predictions is calculated below:
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+#This model will use matrix factorization without tuning parameters
+if(!require(recosystem)) install.packages("recosystem", repos = "http://cran.us.r-project.org")
+set.seed(69, sample.kind = 'Rounding')
+
+train_set <- with(edx, data_memory(user_index=userId, item_index=movieId, rating=rating))
+test_set <- with(validation, data_memory(user_index=userId, item_index=movieId, rating=rating))
+
+# Create a model object
+r <- Reco()
+
+# Tune parameters with cross validation
+opts <- r$tune(train_set, opts = list(dim=c(10, 20, 30), #dim is number of latent factors
+                                      lrate=c(0.1, 0.2), #learning rate (step size in gradient decent)
+                                      costp_l1=0, costq_l1=0, #L1 regularization terms set to 0 (default is c(0.01, 0.1))
+                                      nthread=1, #number of threads for parallel computing
+                                      niter=10)) #number of iterations
+#train with tuned parameters
+r$train(train_set, opts=c(opts$min, nthread=1, niter=20))
+
+#predict
+final_prediction <- r$predict(test_set, out_memory())
+
+(final_rmse <- RMSE(final_prediction, validation$rating))
+```
 
@@
 Verifying that the updating rmse is behaving as expected is an easy way of making sure the model has not broken down.
+
+# Citation
+
+F. Maxwell Harper and Joseph A. Konstan. 2015. The MovieLens Datasets: History and Context. ACM Transactions on Interactive Intelligent Systems
+(TiiS) 5, 4, Article 19 (December 2015), 19 pages. DOI=http://dx.doi.org/10.1145/2827872

### 2021-03-21 main.R

@@ -1,3 +1,4 @@
+# Ray Sherbourne
 ##########################################################
 # Create edx set, validation set (final hold-out test set)
 ##########################################################

### 2021-03-22 draft4.Rmd
touchups

@@
 
-To follow up on the previous idea of using time as a predictor.
+The following bar chart shows the number frequency of movie genres. Drama and Comedy stand out as the biggest genres. 
+```{r, echo=FALSE}
+(genre_count <-edx %>% distinct(edx$movieId, .keep_all = TRUE) %>% select(9:26) %>% 
+    colSums()) %>% barplot(las=2, main = 'Movie Genres', ylab = 'Count')
+```
+
+To follow up on a previous idea of using time as a predictor.
 A check is done using movies containing the word 'Christmas' to see if holiday themed movies should be recommended more around that holiday.
@@
 Cross-validation is used to search for the optimal lambda term.
+A plot of the lambda search is shown below:
 
@@
 # save the best lambda term
-(lambda <- lambdas[which.min(rmses)])
+lambda <- lambdas[which.min(rmses)]
 
@@
 In tuning the parameters, multiple values were experimented with, however this process took a significant amount of computational time, so only these two models are included here.
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
+```{r, include=FALSE}
 #This model will use matrix factorization without tuning parameters
@@
 matrix_factor_tuned_rmse <- RMSE(matrix_factor_tuned, edx_test$rating)
+```
 
+```{r, echo=FALSE, message=FALSE}
 # Evaluate the different models
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
+```{r, include=FALSE}
 #This model will use matrix factorization without tuning parameters
@@
 final_prediction <- r$predict(test_set, out_memory())
+```
 
+```{r, echo=FALSE, message=FALSE}
 (final_rmse <- RMSE(final_prediction, validation$rating))

### 2021-03-22 main.R

@@ -92,8 +92,7 @@ edx <- edx %>% mutate(Action = ifelse(str_detect(.$genres, 'Action'), 1, 0),
                       Thriller = ifelse(str_detect(.$genres, 'Thriller'), 1, 0),
                       War = ifelse(str_detect(.$genres, 'War'), 1, 0),
                       Western = ifelse(str_detect(.$genres, 'Western'), 1, 0))
-(genre_count <-edx %>% distinct(edx$movieId, .keep_all = TRUE) %>% select(9:26) %>% 
-  colSums()) %>% barplot(las=2, main = 'Movie Genres', ylab = 'Count')
+
 # Visualize the predictors
 
 # User ratings
@@ -114,7 +113,8 @@ edx %>% mutate(date=round_date(date, unit='week')) %>% group_by(date) %>%
   ggplot(aes(x=date, y=rating)) + geom_point() + geom_smooth()+
   ggtitle('Average Rating per week') + ylab('Rating')
 # Genres
-(edx %>% group_by(movieId) %>% pull() %>% colSums(edx[,9:26]))
+(genre_count <-edx %>% distinct(edx$movieId, .keep_all = TRUE) %>% select(9:26) %>% 
+    colSums()) %>% barplot(las=2, main = 'Movie Genres', ylab = 'Count')
 
 # Christmas movies (proof that holiday movie discrimination is pointless)
 christmas_pattern <- "Christmas"

### 2021-03-23 Sherbourne_Capstone.Rmd
Capstone

--- 2021-03-22 draft4.Rmd
@@
 The genres variable contains a combinations of 18 possible genres.
-Looking at the first few lines of data, there are a few data munging operations that will help with further processing:
+Below is a look at the first few lines of data:
 
@@
 ```
+
+There are a few data munging operations that will help with further processing:
 
@@
 
-The following bar chart shows the number frequency of movie genres. Drama and Comedy stand out as the biggest genres. 
+The following bar chart shows the number of movie genres. Drama and Comedy stand out as the biggest genres. 
 ```{r, echo=FALSE}
@@
 ```{r, echo=FALSE, message=FALSE}
-(final_rmse <- RMSE(final_prediction, validation$rating))
+final_rmse <- RMSE(final_prediction, validation$rating)
+cat('Final RMSE: ', final_rmse)
 ```

### 2021-03-23 Sherbourne_Capstone.R

@@ -1,0 +1,302 @@
+# Ray Sherbourne
+##########################################################
+# Create edx set, validation set (final hold-out test set)
+##########################################################
+
+# Note: this process could take a couple of minutes
+
+if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
+if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
+if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
+
+library(tidyverse)
+library(caret)
+library(data.table)
+
+# MovieLens 10M dataset:
+# https://grouplens.org/datasets/movielens/10m/
+# http://files.grouplens.org/datasets/movielens/ml-10m.zip
+
+dl <- tempfile()
+download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
+
+ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
+                 col.names = c("userId", "movieId", "rating", "timestamp"))
+
+movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
+colnames(movies) <- c("movieId", "title", "genres")
+
+# if using R 3.6 or earlier:
+#movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
+#title = as.character(title),
+#genres = as.character(genres))
+# if using R 4.0 or later:
+movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
+                                           title = as.character(title),
+                                           genres = as.character(genres))
+
+
+movielens <- left_join(ratings, movies, by = "movieId")
+
+# Validation set will be 10% of MovieLens data
+set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
+test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
+edx <- movielens[-test_index,]
+temp <- movielens[test_index,]
+
+# Make sure userId and movieId in validation set are also in edx set
+validation <- temp %>% 
+  semi_join(edx, by = "movieId") %>%
[…]

### 2021-03-23 main.R
Delete main.R

@@ -1,302 +1,0 @@
-# Ray Sherbourne
-##########################################################
-# Create edx set, validation set (final hold-out test set)
-##########################################################
-
-# Note: this process could take a couple of minutes
-
-if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
-if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
-if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
-
-library(tidyverse)
-library(caret)
-library(data.table)
-
-# MovieLens 10M dataset:
-# https://grouplens.org/datasets/movielens/10m/
-# http://files.grouplens.org/datasets/movielens/ml-10m.zip
-
-dl <- tempfile()
-download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
-
-ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
-                 col.names = c("userId", "movieId", "rating", "timestamp"))
-
-movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
-colnames(movies) <- c("movieId", "title", "genres")
-
-# if using R 3.6 or earlier:
-#movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
-#title = as.character(title),
-#genres = as.character(genres))
-# if using R 4.0 or later:
-movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
-                                           title = as.character(title),
-                                           genres = as.character(genres))
-
-
-movielens <- left_join(ratings, movies, by = "movieId")
-
-# Validation set will be 10% of MovieLens data
-set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
-test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
-edx <- movielens[-test_index,]
-temp <- movielens[test_index,]
-
-# Make sure userId and movieId in validation set are also in edx set
-validation <- temp %>% 
-  semi_join(edx, by = "movieId") %>%
[…]
